Solar-Electric and Gas Powered, Long-Endurance UAV Sizing via Geometric Programming

### 2017-04-25 docs/reviews/review_response.tex

We thank the anonymous referees for their thoughtful and insightful comments on the draft manuscript.
We feel that the paper has improved significantly thanks to the feedback. 
Specific comments from the reviews (in italics) are addressed herein.

\section*{Referee 1}

\emph{
This article presents a description of a methodology to design a hale type aircraft using geometric programming.
As such, the approach is probably of more relevance than the specifics.
It lacks focus. Frankly this article is far more appropriate as a conference paper than a journal article. It reads like a recipe with a significant number of assumptions that are not supported.
While the approach is definitely of value, the presentation is less so. If published, this article should be shortened to an engineering note. }

We thank the referee for this comment. 
We have made organizational changes to the paper, including a more detailed description of the requirements, GP approach, and results. 
As with any aircraft design, assumptions had to be made.  
We did our best to explicity state and justify major assumptions about the requirements.   
We would be happy to respond to any sepecific assumptions that the referee feels are not supported. 
We would like to point out that justifying all of our assumptions would significantly lengthen the paper.
We also aim to show that because of the GP approach we can quickly change our assumptions and resolve. 
We do this in the results section of the paper and believe this shows important insights about long-endurance UAVs.  \\

@@

We thank the referee for this comment.  We did cut a few unnecessary equations to help shorten the paper. \\

\section*{Referee 2}

\emph{Optimization: I would like to see more information on what were the design variables for both aircraft, how long a single analysis took and how many analyses were required by the GP approach.}

We would like to thank the referee for this suggestion.  
We claim that speed is o of great benefits of using GP and we should have included more information about our design variables, solve times, and number of analysis.  
We refer the referee to the results section of the revised paper where we have included a table of key design parameters and solve times and number of analysis throughout.  \\

@@

\emph{Approximation: In order to accommodate the use of GP posynomial approximations were used. It is important then to check the final designs and see if they perform as approximated. Often the optimization process is steered to regions where the quality of the approximations is poor.}

We thank the referee for this insightful point.  
We have since included bounds on each posynmial approxmiation to ensure that the solution does not enter a region where we do not have representative data.  
We noticed when setting these bounds that the solar-electric solution tended to operate a lower Reynolds numbers than was approximated by our posynomial approxmiation.  
We have since generated more drag data from XFOIL at low Reynolds numbers and re-generated the posynomial fits for the airfoil drag data. 
We also implemented a post-solution ``check''  for any posynomial fit that relies on XFOIL data.  
Once a solution has been found the optimized lift coefficient and Reynolds number are entered into XFOIL and the drag coefficient is compared.  
We guarentee that all drag coefficients are within 5\% of the predicted XFOIL drag number.\\

@@

\emph{Assumptions:
The paper appears to insist that a single model of the UAV need to be used for all latitudes. This seems to be very restrictive. I would like to know how much better the performance is for UAVs that are tailored for a narrow band of latitudes.
In a similar vein, why not have both kind of UAVs depending on the latitude and the season.
}

We thank the referee for this comment.  These are both interesting trade studies that we could certainly explore.  We chose not to include them in this paper because we felt they were not central take aways of our findings.  This would have additionally increased the length of our paper.  If the referee feels that these two trade studies should be explored and included we would be happy to discuss this futher. \\

\emph{Length: The paper is tedious to read with the long listing of models and constraints. It may make more sense to put some of the material as additional on-line material, which I assume is available with the Journal of Aircraft.}

We recognize that this paper is long.  
We have cut a few non-essential equations and graphs to try and shorten the paper.  
However, because GP solves a list of constraints we felt it important to explicity state all of our equations to justify our results.  
We are willing to discuss cutting additionally sections/figures/equations if the referee has specific suggestions. \\

\emph{Minor issues}
\begin{itemize}
    \item \emph{The fact that the design is limited to the northern hemisphere should be articulated clearly in the Abstract.} \\
We thank the referee for this comment.  
We recongnize that this may have seemed like a limitation in our assumptions.  
We have since revised our study to include wind speed data from the Southern Hemisphere.  
We would refer the referee to Figure 1 of the revised paper.  
We initially chose not to include wind speed data from the Southern Hemisphere because the wind speeds are consistently lower in the Southern Hemisphere versus the Northern Hemisphere.  
Thus our assumption was conservative.  
However, given that our results do not change drastically by including the Southern Hemisphere wind speed data we have chosen to included it to avoid misconceptions about our results. \\

    \item \emph{Sources for data presented in figures should be given in the captions.} \\
        We thank the reader for this comment.  Citations have since been added to all capations where data is used.  \\

@@

    \item \emph{It was not clear to me why the solar UAV not fly substantially higher as there is less drag even if wind speeds are higher.} \\
    There are two separate, but related issues that affect the operating altitude.  The first is wind speed.  It is true that drag is lower at higher altitudes.  But, if we want to be faster than the wind speeds then we pay a higher penalty for wind speeds than we do for air density,

    \begin{equation}
        D = \frac{1}{2} \rho V^2 S C_D
    \end{equation}

    as drag goes by $V^2$ but only varies linearly with $\rho$.  Additionally, wind speeds cause an even greater constraint on power because power goes by velcoity cubed, $P \sim V^3$.  This affects the number of solar cells and number of batteries, which in turn affects the wing size and therefore aircraft weight. This would suggest that we should fly at the minimum wind speed always.

    There is a second factor that is also going on, which is the trade off between wing weight and air density.  Because lift also scales equally with air density and wing area,

@@

    \begin{equation}
        L = \frac{1}{2} \rho V^2 S C_L
    \end{equation}

    for lower air densities a bigger wing is required.  Because of the cubed square law, a larger wing is very costly.  Therefore, it is actually more advantageous to fly at a lower altitude to accomodate a slightly larger wing.

    If we assume that a large wing doesn't affect the weight at all, then we should see the aircraft fly at the minimum wind speed.  This is precisely the point we are trying to make in Figure 17.  \\

    \item \emph{I did not understand what the autocorrelation indicated in terms of performance. That is, I could not make sense of the meaning of Figs 4a and 5a.}  \\
We thank the referee for this comment.  We realized that these autocorrelation plots, while interesting, are not actually included to the optimization model and we chose to cut them from the paper.  
We initially included them because as we were doing the research we wanted to understand what a percentile wind speed meant operationally.  
We were asking ourselves does the 90th percentile wind speed mean that everyday we will be blown off station 10\% of the time, or that there will be 5 days in the month during which it is impossible to station keep.  
The autocorrelation plots showed that the later was true.  In other words, when there are high wind speeds they can occur for days a time. \\

    \item \emph{When data is given for a certain latitude, it is not clear whether the implication is that there is no difference with longitude, or that this is the worst for any longitude.} \\
While wind speeds certianly vary with longitude, our claim is that the variation with longitude is minimial compared to the variation to latitude and can be ignored.  \\

@@

    We thank the referee for this comment.  We realize that we may have been unclear in our comparison.  What we are really comparing is the capability of gas and solar-electric powered aircraft to meet the same set of requirements.  We have changed the second section in the paper to convey that statement.  We believe this to be a fair comparison and will hopefully address other comments made by this referee. \\

\emph{Overall I only had a few other comments:}

\begin{itemize}
    \item \emph{On page 5: again this speaks to not having the same mission constraints for both aircraft. Using the winter solstice to set the endurance level for the solar aircraft does not truly examine the design space for that aircraft. From a mission perspective having an aircraft that can fly 6 months of the year can be desirable when you are comparing it to an aircraft with a 7 day endurance. Even though that 7 day endurance can happen at any time during the year.} \\

        We believe this comment is best addressed by a clarification of the requirements.  One requirement states that the mission must be greater than 6 days.  Obviously, the solar-electric powered aircraft will likely exceed 6 days with ease, but it is same requiement for both aircraft.  Another requirement is that the aircraft be able to fly in all seasons.  This requires the solar-electric aircraft to fly during the winter-solstice.  As we explored changing the latitude requirement, we could also explore changing the season requirement.  We chose not to explore this trade study in the paper because of the additional length to the paper.  We would be happy to discuss includind a season trade study if the referee feels it is an important result. \\

@@

    \item \emph{On page 8: it should be stated that the temperature is constant between the altitude range of 11 km up to 20 km not above 11 km. Also figure 3 shows the altitude in ft. This is not consistent with the text which states altitude in meters. A single unit system should be used, otherwise it gets confusing. I would suggest listing everything in meters and then if needed put the equivalent value in ft in parenthesis next to it.} \\

        We thank the referee for noticing this error.  It has since been corrected.  Additionally, all altitude references in the paper are now in meters.\\

@@

    \item \emph{On page 12: in figure 6 it should be indicated that the markers "o" are data points for the wind speed and their source should be referenced on the figure.} \\

        We thank the referee for noticing this.  The data points have since been labeled and their source referenced.  \\

    \item \emph{On page 26: Using the fuselage to store all of the fuel for the combustion engine aircraft and then using the wings to store the batteries for the solar electric aircraft is not consistent between the designs. Again, this is fine for just demonstrating the optimization method but is an issue if you are trying to compare the results between the two aircraft types and draw a conclusion.} \\

        We thank the referee for the insightful comment.  Because we are attempting to do a fair comparison between the two archictures we have put a lot of thought into whether or not to putting fuel in the wings is the best possible configuration for the gas powered archicture.  We could certainly add it as a constraint to the optimization but to be fair we would have to add additional constraints to account for the added weight and complexity of wet wings.  For the smaller gas powered UAVs that are being considered in this problem wet wings would add a significant amount of weight and complexity whereas a fuselage of this size is easy to build and light weight.  
        For the solar-electric aircraft putting batteries in the wings is much simpler than putting fuel in wings.  
        Batteries do not require a special fuel tank to hold liquid and have low volume density.  
        Therefore, we believe that putting batteries in the wings for the solar-electric aircraft and fuel in a fuselage for the gas aircraft is a fair comparison.  \\

### 2017-04-26 docs/paperin.tex
print to eps not pdf where possible

The wind speed at a station is a function of latitude, altitude, season or day of the year, and percentile wind speed,

@@

For long-endurance gas powered aircraft, the endurance requirement is defined by amount of time the aircraft needs to remain in the air without refueling.  
For a solar-electric powered aircraft, however, if it can fly during the shortest day of the year, December 21st (June 21st for Southern Hemisphere), then it can thoretically fly at any time of the year, more than satisfying the 6 day endurance requirement. 
Therefore, to meet the endurance requirement the solar-electric powered aircraft must be able to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section IV.B, and with higher than average wind speeds as shown in Figure~\ref{f:windvsmonth}.  
Because becuase wind speeds are greatest during the winter solstice for both the Northern and Southern Hemispheres only winds speeds from December and June will be used for the aircraft sizing and performance analysis presented in this paper.

\begin{figure}[H]
 \begin{subfigmatrix}{2}% number of columns
     \subfigure[-30$^{\circ}$ latitude \label{f:windvsmonthS30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthS30.pdf}}
     \subfigure[+30$^{\circ}$ latitude \label{f:windvsmonthN30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthN30.pdf}}
 \end{subfigmatrix}
 \caption{\textbf{Winds peak at summer and winter solstice in Southern and Northern Hemispheres respectively.\cite{wind}}}
 \label{f:windvsmonth}
\end{figure}

@@

Long-endurance aircraft tend to fly at low velocities, which based on Figure~\ref{f:altvswindhmin} means there are two possible operating altitude regiems: at or around the minimum payload requirement 4,600 - 7,600 m or at high altitudes between 16,700-20,000 m. 
Gas powered aircraft generally fly in the lower altitude regime of 4,600-7,600 m because naturally aspirated engines lose power with increased altitude and are unable to reach altitudes higher than about 13,500 m depending on the engine.  
Solar-electric powered aircraft, which do not have naturally aspirated engines, will to fly around 18,000 m to avoid cloud coverage.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=538,natheight=405]{altvswindhmin.pdf}
    \caption{\textbf{Aircraft must fly above $h_{\text{min}}$ to meet altitude requirement. Bands represent 80th, 90th and 95th percentiles.\cite{wind}}}
	\label{f:altvswindhmin}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=576,natheight=432]{latvswindN.pdf}
    \caption{\textbf{Wind speeds by latitude.  Bands represent 80th, 90th and 95th percentile winds.\cite{wind} }}
	\label{f:latvswind}
	\end{center}
\end{figure}

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section IV.B.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=480,natheight=200]{flowchart.pdf}
    \caption{\textbf{Process of constructing and solving a GP optimization problem. }}
	\label{f:flowchart}
	\end{center}
\end{figure}

A python package for defining and manipulating geometric programs, GPkit\cite{gpkitdocs}, is used to create the models described in this paper.  
A commercial solver, mosek\cite{mosek}, is used to solve the GP. 
Both the gas and solar-electric powered GP optimization models are available for download and use. \cite{gas_solar_trade} \\

@@

Various structural, aerodynamic, and weight models necesistate assumptions about the vehicle configuration.  
Common characteristics between the two models include a constant tapered wing and a conventional tail with a single tail boom extending from the wing.  
The gas powered aircraft has a fuselage to hold all of the fuel required for the mission.  
The solar-electric powered aircraft holds the batteries in the wings.  
The solar cells for the solar-electric aircraft are placed along the wing and possibly on the horizontal tail.  
A simple diagram of what each vehicle might look like is shown in Figure~\ref{f:simpleaircraft} with their corresponding weight breakdowns in Figure~\ref{f:wbreak}.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=517,natheight=408]{simpleaircraft.pdf}
    \caption{\textbf{Blue and red are specific to the solar-electric and gas archictures respectively.  Black are shared characteristics.}}
	\label{f:simpleaircraft}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.7\textwidth,natwidth=358,natheight=334]{wbreak.pdf}
   \caption{\textbf{Representative weight breakdown of both aircraft optimization models.}}
	\label{f:wbreak}
	\end{center}
\end{figure}

@@

Because wind speeds also vary with latitude, several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg et. al\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density.

For example, an aircraft optimized for $\pm$30 degrees latitude will have 11 different wind speed and solar flux constraints that must be satisfied, one for each degree of latitude between 20 and 30 degrees. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude, no constraints for latitudes below 20 degrees are actually included in the optimization. 
The wind speed data used to generate these equations includes December wind speeds from the Northern Hemisphere and June wind speeds from the Southern Hemisphere for the year 2005-2015.
In total, 41 wind speed constraints were generated to represent $\pm$20-60 degrees. (See Appendix A) 
The constraint on December wind speeds and air density at the 30 degree latitude is

\begin{align}
    \left(\frac{V_{\text{wind}}}{V_{\text{ref}}}\right)^{7.066} &\geq 18413.2\rho^{9.3}p_{\text{wind}}^{5.51} + 9658.5\rho^{8.25}p_{\text{wind}}^{23.47} \nonumber \\
    \label{e:windfit35}
    &+ \num{2.55d-13}\rho^{-7.72}p_{\text{wind}}^{49.65} + \num{1.11d-16}\rho^{-9.80}p_{\text{wind}}^{10.26} + 10.493.8\rho^{7.31}p_{\text{wind}}^{226.9}
\end{align}

where $V_{\text{ref}} = 100$ [m/s] and $p_{\text{wind}}$ is the percentile wind speed. The comparison of the fitted Equation~\eqref{e:windfit35} to the actual data is presented in Figure~\ref{f:windfitl35}.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=511,natheight=405]{windfitl30.pdf}
    \caption{\textbf{Posynomial approximation to wind speed data\cite{wind} (circles) at 35$^{\circ}$ latitude. (Log-space RMS error = 0.043.)}}
	\label{f:windfitl35}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.7\textwidth,natwidth=572,natheight=405]{lat30.pdf}
    \caption{\textbf{Solar power on Dec 21st at 30$^{\circ}$ North. 24-hour operationality achieved when red area exceeds blue, magenta and green areas. }}
	\label{f:lat30}
	\end{center}
\end{figure}

@@

where $\eta_{\text{motor}} = 0.95$. Thus for a given latitude and day, and therfore total solar energy $(E/S)_{\text{sun}}$, both $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ are functions of $(P/S)_{\text{min}}$.  
The total solar energy available $(E/S)_{\text{sun}}$, is computed prior to the optimization solve and is used to generate monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ as functions $(P/S)_{\text{min}}$, one for each degree of latitude between 20 and 60 degrees latitude. 
% Figure~\ref{f:energyapprox} shows sample monomial approximations at 30 degrees latitude. 
% 
% \begin{figure}[H]
%  \begin{subfigmatrix}{2}% number of columns
%      \subfigure[Monomial Approxmiation for $(E/S)_{\text{day}}$\label{f:Benergy}]{\includegraphics[natwidth=576,natheight=432]{Benergy.pdf}}
%      \subfigure[Monomial Approximation for $(E/S)_{\text{twilight}}$\label{f:Cenergy}]{\includegraphics[natwidth=575,natheight=432]{Cenergy.pdf}}
%  \end{subfigmatrix}
%  \caption{\textbf{Monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$. (Log-space RMS error = 0.007 and 0.0180)}}
%  \label{f:energyapprox}
% \end{figure}

@@

    Using a trapezoidal integration of Equation~\eqref{e:solares}, the total available solar energy per unit area can be obtained for a given latitude and day of the year. Because only Equations~\eqref{e:solarreq}-\eqref{e:solarssolar} are GP compatible, the total solar energy per unit area per day, the length of the day and the length of the night are calculated from the latitude and the day of the year prior to an optimization solve.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=524,natheight=405]{powervsweightfit.pdf}
    \caption{\textbf{Power law fit to University of North Dakota engine weight data\cite{gasengine}. (Log-space RMS error = 0.34)}}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=530,natheight=414]{powertobsfcfit.pdf}
    \caption{\textbf{Representative engine performance fit based on RCV Engine Ltd data.  (Log-space RMS error = 0.007) }}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

@@

    To estimate the wing profile drag, a posynomial fit is made of the drag polar of the selected \emph{JH01} airfoil (See Appendix B for a discussion of the airfoil choice). 
    Drag polars were produced using XFOIL\cite{xfoil} at various Reynolds and the data was then fit to a posynomial equation using techniques described in Hoburg et. al.\cite{fitting}
    The XFOIL drag polar data is compared to the posynomial fit in Figure~\ref{f:JH01polar}.

    \begin{align}
        \label{e:aerodragprof}
        c_{d_p}^{3.72} &\geq 0.0247C_L^{2.49}Re^{-1.11} + 2.03e^{-7}C_L^{12.7}Re^{-0.338} + 6.35e^{10}C_L^{-0.243}Re^{-3.43} \nonumber \\
                       &+ 6.49e^{-6}C_L^{-1.9}Re^{-0.681} \\
        Re &= \frac{\rho V S/b}{\mu}
    \end{align}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=536,natheight=405]{jho1polarfit1.pdf}
    \caption{\textbf{Posynomial fit (solid lines) to XFOIL data (circles).  (Log-space RMS error~=~0.00489.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

@@

\subsubsection{Standard Wing Bending Case}

@@

\subsubsection{Gust Loading Case}

@@

\subsubsection{Discretized Beam Model}

@@

\subsubsection{Cap Spar for Bending Loads}

A cap spar is considered for the solar-electric and gas powered aircraft.  Both aircraft optimization models share the same governing equations.  A cap spar has two carbon fiber caps separated by a foam core as seen in Figure~\ref{f:capspar}. A thin shear web is wrapped around the caps and foam to prevent shearing and buckling.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=3242,natheight=1195]{capsparcross.pdf}
    \caption{\textbf{Cross sectional view of a cap spar.}}
	\label{f:capspar}
	\end{center}
\end{figure}

@@

where $\rho_{\text{cfrp}} = 1.6$ [g/cm$^3$].\cite{carbonfiber}

\subsection{Wing Skin Weight}

@@

where $\rho_{A_{\text{cfrp}}} = 0.049$ [g/cm$^2$], or approximately the area density of one ply of carbon fiber.\cite{carbonfiber} The wing skin is asssumed not to contribute to the bending stiffness.

\subsection{Empennage}

@@

Important trade offs between the gas powered and solar-electric powered archictures are highlighted and quantified by changing the latitude and endurance endurance requirements.   
Both optimization models were solved by minimizing the max take off weight across different bands of latitudes. 
Figure~\ref{f:latvsmtowtrade} shows this result evaluated at the 80th, 90th and 95th percentile wind speeds.  
The gas powered model was solved x times and the solar-electric powered model was solved x times to produce Figure~\ref{f:latvsmtowtrade}.
The gas powered optimization model contains 606 free variables and the solar-electric powered optmization model contains between 165 and 585 free variables depending on latitude band requirement.

>>>>

Observations in wind speed data\cite{wind} indicate that it is unlikely for long-endurance UAVs to be able to station keep in 100 percentile wind speeds.  
Therefore, the station keeping requirement is defined as the percentage of time the aircraft can remain on station, which is the percentile wind speed, $p_{\text{wind}}$. 
The wind speed at a station is also a function of latitude, altitude, season or day of the year,

@@

For long-endurance gas powered aircraft, the endurance requirement is defined by amount of time the aircraft needs to remain in the air without refueling.  
For a solar-electric powered aircraft, however, if it can fly during the shortest day of the year, December 21st (June 21st for Southern Hemisphere), then it can thoretically fly at any time of the year, more than satisfying the 6 day endurance requirement. 
Therefore, to meet the endurance requirement the solar-electric powered aircraft must be able to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section IV.A, and with higher than average wind speeds as shown in Figure~\ref{f:windvsmonth}.  
Because becuase wind speeds are greatest during the winter solstice for both the Northern and Southern Hemispheres only winds speeds from December and June respectively, will be used for the aircraft sizing and performance analysis presented in this paper.

\begin{figure}[H]
 \begin{subfigmatrix}{2}% number of columns
     \subfigure[-30$^{\circ}$ latitude \label{f:windvsmonthS30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthS30.eps}}
     \subfigure[+30$^{\circ}$ latitude \label{f:windvsmonthN30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthN30.eps}}
 \end{subfigmatrix}
 \caption{\textbf{Winds peak at the winter solstice in Southern and Northern Hemispheres.\cite{wind}}}
 \label{f:windvsmonth}
\end{figure}

@@

Long-endurance aircraft tend to fly at low velocities, which based on Figure~\ref{f:altvswindhmin} means there are two possible operating altitude regiems: at or around the minimum payload requirement 4,600 - 7,600 m or at high altitudes between 16,700-20,000 m. 
Gas powered aircraft generally fly in the lower altitude regime of 4,600-7,600 m because naturally aspirated engines lose power with increased altitude and are unable to reach altitudes higher than about 13,500 m.  
Solar-electric powered aircraft, which do not have naturally aspirated engines, will to fly around 18,000 m to avoid cloud coverage.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=538,natheight=405]{altvswindhmin.eps}
    \caption{\textbf{Aircraft must fly above $h_{\text{min}}$ to meet altitude requirement. Bands represent 80th, 90th and 95th percentiles.\cite{wind}}}
	\label{f:altvswindhmin}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=576,natheight=432]{latvswindN.eps}
    \caption{\textbf{Wind speeds by latitude.  Bands represent 80th, 90th and 95th percentile winds.\cite{wind} }}
	\label{f:latvswind}
	\end{center}
\end{figure}

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section IV.A.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=568,natheight=154]{flowchart.eps}
    \caption{\textbf{Process of constructing and solving a GP optimization problem. }}
	\label{f:flowchart}
	\end{center}
\end{figure}

A python package for defining and manipulating geometric programs, GPkit\cite{gpkitdocs}, is used to create the models described in this paper.  
A commercial solver, mosek\cite{mosek}, is used to solve the GP. 
Both the gas and solar-electric powered GP optimization models are available for download and use. \cite{gassolartrade} \\

@@

Various structural, aerodynamic, and weight models necesistate assumptions about the vehicle configuration.  
Common characteristics between the two models include a constant tapered wing and a conventional tail with a single tail boom extending from the wing.  
The gas powered aircraft has a fuselage to hold all of the fuel required for the mission.  
The solar-electric powered aircraft holds the batteries in the wings.  
The solar cells for the solar-electric aircraft are placed along the wing and possibly on the horizontal tail.  
A simple diagram of what each vehicle might look like is shown in Figure~\ref{f:simpleaircraft} with their corresponding weight breakdowns in Figure~\ref{f:wbreak}.
The configuration is assumed to be static for this study.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.9\textwidth,natwidth=447,natheight=265]{simpleaircraft.eps}
    \caption{\textbf{Blue and red are specific to the solar-electric and gas archictures respectively.  Black are shared characteristics.}}
	\label{f:simpleaircraft}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=378,natheight=335]{wbreak.eps}
   \caption{\textbf{Representative weight breakdown of both aircraft optimization models.}}
	\label{f:wbreak}
	\end{center}
\end{figure}

@@

Because wind speeds also vary with latitude, several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg et. al\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. (See Appendix A)

For example, an aircraft optimized for $\pm$30 degrees latitude will have 11 different wind speed and solar flux constraints that must be satisfied, one for each degree of latitude between 20 and 30 degrees. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude, no constraints for latitudes below 20 degrees are actually included in the optimization. 
The wind speed data used to generate these equations includes December wind speeds from the Northern Hemisphere and June wind speeds from the Southern Hemisphere for the years 2005-2015.
In total, 41 wind speed constraints were generated to represent $\pm$20-60 degrees each with a log space RMS error of less than 5\%. (See Appendix B) 
% The constraint on December wind speeds and air density at the 30 degree latitude is 
% 
% \begin{align}
%     \left(\frac{V_{\text{wind}}}{V_{\text{ref}}}\right)^{7.066} &\geq 18413.2\rho^{9.3}p_{\text{wind}}^{5.51} + 9658.5\rho^{8.25}p_{\text{wind}}^{23.47} \nonumber \\
%     \label{e:windfit35}
%     &+ \num{2.55d-13}\rho^{-7.72}p_{\text{wind}}^{49.65} + \num{1.11d-16}\rho^{-9.80}p_{\text{wind}}^{10.26} + 10.493.8\rho^{7.31}p_{\text{wind}}^{226.9}
% \end{align}
% 
% where $V_{\text{ref}} = 100$ [m/s] and $p_{\text{wind}}$ is the percentile wind speed. The comparison of the fitted Equation~\eqref{e:windfit35} to the actual data is presented in Figure~\ref{f:windfitl35}. 
% 
% \begin{figure}[H]
% 	\begin{center}
% 	\includegraphics[width=0.6\textwidth,natwidth=511,natheight=405]{windfitl30.pdf}
%     \caption{\textbf{Posynomial approximation to wind speed data\cite{wind} (circles) at 35$^{\circ}$ latitude. (Log-space RMS error = 0.043.)}}
% 	\label{f:windfitl35}
% 	\end{center}
% \end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.7\textwidth,natwidth=572,natheight=405]{lat30.eps}
    \caption{\textbf{Solar power on Dec 21st at 30$^{\circ}$ North. 24-hour operationality achieved when red area exceeds blue, magenta and green areas. }}
	\label{f:lat30}
	\end{center}
\end{figure}

@@

where $\eta_{\text{motor}} = 0.95$. 
Thus, for a given latitude and day, and therfore total solar energy $(E/S)_{\text{sun}}$, both $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ are functions of $(P/S)_{\text{min}}$.  
The total solar energy available $(E/S)_{\text{sun}}$, is computed prior to the optimization solve and is used to generate monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ as functions $(P/S)_{\text{min}}$, one for each degree of latitude between 20 and 60 degrees latitude. 
$(E/S)_{\text{sun}}$, $(E/S)_{\text{day}}$, $(E/S)_{\text{twilight}}$, and $t_{\text{night}}$ are precomputed using the formulas in Appendix C to generate the fitted equations. 
% Figure~\ref{f:energyapprox} shows sample monomial approximations at 30 degrees latitude. 
% 
% \begin{figure}[H]
%  \begin{subfigmatrix}{2}% number of columns
%      \subfigure[Monomial Approxmiation for $(E/S)_{\text{day}}$\label{f:Benergy}]{\includegraphics[natwidth=576,natheight=432]{Benergy.pdf}}
%      \subfigure[Monomial Approximation for $(E/S)_{\text{twilight}}$\label{f:Cenergy}]{\includegraphics[natwidth=575,natheight=432]{Cenergy.pdf}}
%  \end{subfigmatrix}
%  \caption{\textbf{Monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$. (Log-space RMS error = 0.007 and 0.0180)}}
%  \label{f:energyapprox}
% \end{figure}

@@

\subsubsection{Motor Weight}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=524,natheight=405]{powervsweightfit.eps}
    \caption{\textbf{Power law fit to University of North Dakota engine weight data\cite{gasengine}. (Log-space RMS error = 0.34)}}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=530,natheight=414]{powertobsfcfit.eps}
    \caption{\textbf{Representative engine performance fit based on RCV Engine Ltd data.  (Log-space RMS error = 0.007) }}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

@@

    To estimate the wing profile drag, a posynomial fit is made of the drag polar of the selected \emph{JH01} airfoil (See Appendix D for a discussion of the airfoil choice). 
    Drag polars were produced using XFOIL\cite{xfoil} at various Reynolds and the data was then fit to a posynomial equation using techniques described in Hoburg et. al.\cite{fitting}
    The XFOIL drag polar data is compared to the posynomial fit in Figure~\ref{f:JH01polar}.

%     \begin{align}
%         \label{e:aerodragprof}
%         c_{d_p}^{3.72} &\geq 0.0247C_L^{2.49}Re^{-1.11} + 2.03e^{-7}C_L^{12.7}Re^{-0.338} + 6.35e^{10}C_L^{-0.243}Re^{-3.43} \nonumber \\
%                        &+ 6.49e^{-6}C_L^{-1.9}Re^{-0.681} \\
%         Re &= \frac{\rho V S/b}{\mu}
%     \end{align}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=536,natheight=405]{jho1polarfit1.eps}
    \caption{\textbf{Posynomial fit (solid lines) to XFOIL data (circles).  (Log-space RMS error~=~0.00489.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

@@

\begin{itemize}

\item{Standard Wing Bending Case}

@@

\item{Gust Loading Case}

@@

\item{Discretized Beam Model}

@@

\item{Cap Spar for Bending Loads}

A cap spar is considered for the solar-electric and gas powered aircraft.  A cap spar has two carbon fiber caps separated by a foam core as seen in Figure~\ref{f:capspar}. A thin shear web is wrapped around the caps and foam to prevent shearing and buckling.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.9\textwidth,natwidth=547,natheight=74]{capsparcross.eps}
    \caption{\textbf{Cross sectional view of a cap spar.}}
	\label{f:capspar}
	\end{center}
\end{figure}

@@

where $\rho_{\text{cfrp}} = 1.6$ [g/cm$^3$].\cite{cfply}

\end{itemize}

\subsubsection{Additional Wing Weight}

@@

where $\rho_{A_{\text{cfrp}}} = 0.049$ [g/cm$^2$], or approximately the area density of one ply of carbon fiber.\cite{cfply} The wing skin is asssumed not to contribute to the bending stiffness.

Additional wing weight, $W_{\text{fadd}}$, accounts for additional structural weight in the wing,

\begin{equation}
    W_{\text{fadd}} \geq (W_{\text{spar}} + W_{\text{skin}}) m_{\text{fac}}
\end{equation}

where $m_{\text{fac}} = 1.2$.

\subsubsection{Empennage}

@@

Important trade offs between the gas powered and solar-electric powered archictures are highlighted and quantified by changing the latitude and endurance endurance requirements.   
Both optimization models were solved by minimizing the max take off weight across different bands of latitudes. 
Figure~\ref{f:latvsmtowtrade} shows this result evaluated at the 80th, 90th and 95th percentile wind speeds.  
The gas powered model was solved 63 times in 5.6178 seconds and the solar-electric powered model was solved 31 times in 2.5529 seconds to produce Figure~\ref{f:latvsmtowtrade}.
